---
title: "Fun with ggplot"
author: "Sumeet Kukreja"
date: "9/9/2018"
output: html_document
---

#***********************************************************************************************************
# This code is a comprehensive run throigh for building visualizations using the ggplot2 library 
# Submitted By : Sumeet Kukreja
# emailid : skukre3@uic.edu | sumeetkukreja92@gmail.com | linkedin.com/sumeetkukreja
#***********************************************************************************************************

#******************************************************
# Understanding the Syntax
#******************************************************
#setup
#The most important thing to know while working with ggplot2 is that is only works with a dataframe and not #lists. Also, you can add more layers and themes on an existing ggplot to enhance it accordingly

```{r}
options(scipen=999)  # turn off scientific notation like 2e+08
library(ggplot2)
midwest = data("midwest", package = "ggplot2")  # load the data
midwest <- read.csv("http://goo.gl/G1K41K") # alt source 
head(midwest)
```


```{r}
# Initial Ggplot
ggplot(midwest, aes(x=area, y=poptotal))  # area and poptotal are columns in 'midwest'

```
There is no data points becuase I did not specify if I want a scatter plot or a line chart. Any information that is part of the source dataframe has to be specified inside the aes() function.

```{r}
library(ggplot2)

# Base plot
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state), size=3) +  # Set color to vary based on state categories.
  geom_smooth(method="lm", col="firebrick", size=2) + 
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")

# Change breaks
gg + scale_x_continuous(breaks=seq(0, 0.1, 0.01))
```

If you need to reverse the scale, use scale_x_reverse().

```{r}
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state), size=3) +  # Set color to vary based on state categories.
  geom_smooth(method="lm", col="firebrick", size=2) + 
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")

# Reverse X Axis Scale
gg + scale_x_reverse()
```

To write customized texts for axis labels, I have used 2 methods
Method 1: Using sprintf(). (Have formatted it as % in below example) 
Method 2: Using a custom user defined function. (Formatted 1000’s to 1K scale)
Use whichever method feels convenient.
```{r}
# Base Plot
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state), size=3) +  # Set color to vary based on state categories.
  geom_smooth(method="lm", col="firebrick", size=2) + 
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")

# Change Axis Texts
gg + scale_x_continuous(breaks=seq(0, 0.1, 0.01), labels = sprintf("%1.2f%%", seq(0, 0.1, 0.01))) + 
  scale_y_continuous(breaks=seq(0, 1000000, 200000), labels = function(x){paste0(x/1000, 'K')})
```
To Customize the entire theme in one shot using Pre-Built Themes, I have used 2 different methods.
Method 1: Use the theme_set() to set the theme before drawing the ggplot. Note that this setting will affect all future plots. 
Method 2: Draw the ggplot and then add the overall theme setting (eg. theme_bw())

```{r}
# Base plot
gg <- ggplot(midwest, aes(x=area, y=poptotal)) + 
  geom_point(aes(col=state), size=3) +  # Set color to vary based on state categories.
  geom_smooth(method="lm", col="firebrick", size=2) + 
  coord_cartesian(xlim=c(0, 0.1), ylim=c(0, 1000000)) + 
  labs(title="Area Vs Population", subtitle="From midwest dataset", y="Population", x="Area", caption="Midwest Demographics")

gg <- gg + scale_x_continuous(breaks=seq(0, 0.1, 0.01))

# method 1: Using theme_set()
theme_set(theme_classic())  # not run
gg

# method 2: Adding theme Layer itself.
gg + theme_bw() + labs(subtitle="BW Theme")
gg + theme_classic() + labs(subtitle="Classic Theme")
```

More to come..
+Creating advanced visualizations : modifying theme components, manipulating legend, annotations, faceting and custom layouts.

#***********************************************************************************************************